---
title: "Introducing polaRized: An R Package for Estimating Political Attitude Polarization in Complex Survey Data"
author: David Caldwell
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing polaRized: An R Package for Estimating Political Attitude Polarization in Complex Survey Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This package makes it easy to apply various measures of attitude polarization across potentially many ordered ratings scales in complex survey data. The two main functions are `polarize_assoc` and `polarize_distr`, which return key associational and distributional statitistics, respectively. The package also introduces some extensions to the `survey` package, enabling the incorporation of complex design features into the estimation of statistics commonly used to measure attitude polarization:

- `svykurt` calculates kurtosis.
- `svyskew` calculates standadized skewness.
- `svyextremism` calculates the proportion of extreme responses on ordered ratings scales with different lengths.

Additionally, two helper functions make it easier to prepare data prior to calculating polarization: `filter_scale_length` returns ordered ratings scales that meet a user-defined threshold of unique values, which is useful for statistics that require a certain scale length; `spread_pairs` takes a name key and value key and spreads these columns across unique pairs of observations, making it easier to estimate the association between responses on different survey items.

## Installation

The package can be installed direclty from [GitHub](https://github.com/dcaldwellphd/polaRized) using:

```{r, message=FALSE, warning=FALSE, results = "hide", eval=FALSE}
library(devtools)
devtools::install_github("dcaldwellphd/polaRized")
```

Then load its functions and internal dataset using:

```{r}
library(polaRized)
```

## Data prerequisites

Apart from the `svykurt`, `svyskew`, and `svyextremism`, which can be called on any column containing numerical values, the functions in this package expect data to be stacked in a longer format. There should be a single value column or (for associational measures of polarization) pair of value columns containing observations across groups. The built-in data set (`toydata`) comes in this format.

```{r}
data(toydata)
head(toydata)
```
